An Image-based method for phase estimation, gating and temporal super-resolution of \\cardiac ultrasound

### 2017-02-21 paper155.tex
Updated paper and figures

\begin{abstract}
In this paper, we present a novel purely image-based method for the estimation of instantaneous cardiac and respiratory phases from cardiac ultrasound videos and demonstrate its utility in gating and temporal super-resolution. We first use a novel strategy based on inter-frame similarity to transform the complex high-dimensional image sequence into a 1D time series with the same periodicity characteristics.  Next, we use a trend extraction technique to decompose this time series into a trend component encoding respiratory motion alone and a residual component encoding the beating heart motion alone. Next, we use the Hilbert transform to estimate the instantaneous cardiac and respiratory phases of each frame from these components. Next, we present a robust non-parametric regression technique that uses these phase estimates to gate out video frames predominantly influenced by respiratory motion. Lastly, we use a kernel regression method with a novel kernel to learn an image manifold parameterized by instantaneous cardiac phase that can be sampled to generate a single-cycle video at a higher temporal-resolution. We demonstrate and validate our methods using 2D cardiac ultrasound videos and associated ECG recordings of 6 mice.
%\keywords{Cardiac, Ultrasound, Phase estimation, Gating, Temporal super-resolution}
\end{abstract}
%
\vspace{-0.5cm}
\section{Introduction}
\label{sec:intro}
%
Cardiovascular disease is the leading cause of death worldwide and ultrasound is emerging as an increasingly effective tool for rapid non-invasive visual assessment of cardiac structure and function. Cardiac ultrasound videos consists of two kinds of periodic motion, one that corresponds to the beating heart motion and the other that corresponds to respiratory motion. Knowing the phase or location of each video frame within the cardiac and/or respiratory cycles is essential in a variety of applications such as gating, quiescence detection, 3D reconstruction, and robotic beating heart surgery\cite{VonBirgelen1997,Wick2013,Wachinger2012,Kurz2015}. Typically, the position within the cardiac cycle is tracked through ECG data and the position within respiratory cycle is tracked by motion of markers placed on the subject's body\cite{Khamene2004}.

	In this paper, we present a novel method that estimates the instantaneous cardiac and respiratory phases directly from the images of the cardiac ultrasound video. We also present a robust non-parametric regression technique for gating out respiratory frames and a novel kernel regression to reconstruct images at any cardiac phase to facilitate temporal super-resolution.	

@@

	We first compute the similarity between all pairs of frames in the given periodic image sequence containing $N$ images/frames to create a matrix $A \in R^{N \times N}$ where in the element $A(i,j)$ is equal to the similarity between the $i^{th}$ and $j^{th}$ frames of the given sequence. Here in, we use normalization correlation also known as the Pearson correlation coefficient to quantify inter-frame similarity but, in principle, any of the similarity metrics used in image registration algorithms \cite{Goshtasby2012} can potentially be chosen. Each row in the inter-frame similarity matrix $A$ can now be seen as a univariate time series that measures the similarity of the corresponding frame in the image sequence with all other frames. If the similarity metric is chosen with care and if the corresponding frame is not significantly corrupted we can expect this time series to encompass the predominant periodicity characteristics of the original image sequence. Two kinds of periodic motion are present in cardiac ultrasound videos, one corresponding to the beating heart motion and the other corresponding to respiratory motion. Figure \ref{fig:phase_estimation}(a) shows the inter-frame similarity matrix of one of our cardiac ultrasound videos computed using the normalized correlation metric wherein these two sources of periodicity can be clearly observed in many rows/columns.

@@

	We encourage the readers to watch the supplementary videos for a better assessment of the results of the proposed methods.
%
\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\begin{tabular}[c]{@{}c@{}}VID \\ $\;$ \end{tabular} & \begin{tabular}[c]{@{}c@{}}LOOCV (50) NCORR\\ $mean \pm std$\end{tabular} & \begin{tabular}[c]{@{}c@{}}QRS Peak Frame NCORR\\ $mean \pm std$\end{tabular} \\ \hline
1 & 0.8562 $\pm$ 0.0366 & 0.7981 $\pm$ 0.0532 \\ \hline
2 & 0.8558 $\pm$ 0.0525 & 0.7877 $\pm$ 0.0503 \\ \hline
3 & 0.8592 $\pm$ 0.0348 & 0.7849 $\pm$ 0.0457 \\ \hline
4 & 0.8713 $\pm$ 0.0348 & 0.8395 $\pm$ 0.0486 \\ \hline
5 & 0.8655 $\pm$ 0.0312 & 0.8196 $\pm$ 0.0476 \\ \hline
6 & 0.8736 $\pm$ 0.0278 & 0.8270 $\pm$ 0.0778 \\ \hline
\end{tabular}
\caption{Evaluation of our kernel regression model for reconstructing the image at any cardiac phase using 50 rounds of Leave one out cross-validation (LOOCV)}
\label{table:image_reconstruction}
\end{table}
%

%
\vspace{-0.3cm}
\section{Discussion and future work}
\label{sec:discussion}
%

%
\vspace{-0.3cm}
\bibliographystyle{splncs03}
\bibliography{library}
\end{document}

>>>>

\begin{abstract}
Ultrasound is emerging as an increasingly effective tool for rapid non-invasive visual assessment of cardiac structure and function.
Knowing the phase or location of each video frame within the cardiac and/or respiratory cycles is essential in a variety of applications.
In this paper, we present a novel method for the estimation of instantaneous cardiac and respiratory phases directly from cardiac ultrasound videos. The method: (i) transforms the complex high-dimensional image sequence into a univariate time series with the same periodicity characteristics, (ii) decouples the sources of beating heart and respiratory motion using a trend extraction technique, and (iii) estimates the cardiac and respiratory phases using Hilbert transform. We also present a robust non-parametric regression technique for gating out respiratory frames and a novel kernel regression to reconstruct images at any cardiac phase to facilitate temporal super-resolution. We demonstrate and validate our methods using 2D cardiac ultrasound videos and associated ECG recordings of 6 mice.
%\keywords{Cardiac, Ultrasound, Phase estimation, Gating, Temporal super-resolution}
\end{abstract}
%
\vspace{-0.5cm}
\section{Introduction}
\label{sec:intro}
%
Cardiovascular disease is the leading cause of death worldwide and ultrasound is emerging as an increasingly effective tool for rapid non-invasive visual assessment of cardiac structure and function. Cardiac ultrasound videos consists of two kinds of periodic motion, one that corresponds to the beating heart motion and the other that corresponds to respiratory motion. Knowing the phase or location of each video frame within the cardiac and/or respiratory cycles is essential in a variety of applications such as gating, quiescence detection, 3D reconstruction, and robotic beating heart surgery\cite{VonBirgelen1997,Wick2013,Wachinger2012,Kurz2015}. Typically, the position within the cardiac cycle is tracked through ECG data and the position within respiratory cycle is tracked by motion of markers placed on the subject's body\cite{Khamene2004}.

	In this paper, we present a novel method that estimates the instantaneous cardiac and respiratory phases directly from the images of the cardiac ultrasound video and demonstrate its utility in respiratory gating and temporal super-resolution. We also present a robust non-parametric regression technique for gating out respiratory frames and a novel kernel regression to reconstruct images at any cardiac phase to facilitate temporal super-resolution.	

@@

	We first compute the similarity between all pairs of frames in the given periodic image sequence containing $N$ images/frames to create a matrix $A \in R^{N \times N}$ where in the element $A(i,j)$ is equal to the similarity between the $i^{th}$ and $j^{th}$ frames of the given sequence. Here in, we use normalization correlation also known as the Pearson correlation coefficient to quantify inter-frame similarity but, in principle, any of the similarity metrics used in image registration algorithms can potentially be chosen. Each row in the inter-frame similarity matrix $A$ can now be seen as a univariate time series that measures the similarity of the corresponding frame in the image sequence with all other frames. If the similarity metric is chosen with care and if the corresponding frame is not significantly corrupted we can expect this time series to encompass the predominant periodicity characteristics of the original image sequence. Two kinds of periodic motion are present in cardiac ultrasound videos, one corresponding to the beating heart motion and the other corresponding to respiratory motion. Figure \ref{fig:phase_estimation}(a) shows the inter-frame similarity matrix of one of our cardiac ultrasound videos computed using the normalized correlation metric wherein these two sources of periodicity can be clearly observed in many rows/columns.

@@

	Lastly, since this paper deals with video analysis we encourage the readers to review the supplementary material for a better understanding of the utility and potential of the proposed methods.
%
\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|}
\hline
\begin{tabular}[c]{@{}c@{}}VID \\ $\;$ \end{tabular} & \begin{tabular}[c]{@{}c@{}}LOOCV (50) NCORR\\ $mean \pm std$\end{tabular} & \begin{tabular}[c]{@{}c@{}}QRS Peak Frame NCORR\\ $mean \pm std$\end{tabular} \\ \hline
1 & 0.8562 $\pm$ 0.0366 & 0.7981 $\pm$ 0.0532 \\ \hline
2 & 0.8558 $\pm$ 0.0525 & 0.7877 $\pm$ 0.0503 \\ \hline
3 & 0.8592 $\pm$ 0.0348 & 0.7849 $\pm$ 0.0457 \\ \hline
4 & 0.8713 $\pm$ 0.0348 & 0.8395 $\pm$ 0.0486 \\ \hline
5 & 0.8655 $\pm$ 0.0312 & 0.8196 $\pm$ 0.0476 \\ \hline
6 & 0.8736 $\pm$ 0.0278 & 0.8270 $\pm$ 0.0778 \\ \hline
\end{tabular}
\caption{Evaluation of our kernel regression model for reconstructing the image at any cardiac phase using 50 rounds of Leave one out cross-validation (LOOCV)}
\label{table:image_reconstruction}
\end{table}
%
%
\vspace{-0.3cm}
\section{Conclusion}
\label{sec:conclusion}
%
In this paper, we have presented a novel method to estimate the instantaneous cardiac and respiratory phases directly from the cardiac ultrasound video. We have also presented a robust non-parametric regression technique for gating out respiratory frames and a novel kernel regression model for reconstructing images at any cardiac phase that can be used to facilitate temporal super-resolution. Having demonstrated a proof of concept, in the future, we plan to evaluate our methods on more datasets, especially on cardiac ultrasound videos of subjects with irregular heart beats, and address any remaining pitfalls. One drawback of our phase estimation method is a strong assumption of periodicity which may breakdown in the case of subjects with cardiac arrhythmia. To address this, we will look into generalizations of the hilbert transform\cite{Lu2013} that are more susceptible noise. Another drawback is that the use normalized correlation to measure inter-frame similarity relies on an inherent assumption that a large part of the image is pulsating. To address this, we will try to devise more sophisticated patch-based similarity measures.
%
\vspace{-0.3cm}
\bibliographystyle{splncs03}
\bibliography{library}
\end{document}
